Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-12-04 paper/main.tex
rewrite introductory phrase of application 2

\begin{figure}[H]
    \centering
    \begin{tikzonimage}[width=0.3\textwidth]
        {figures/crm_amir/progressions/step4-000010.png}
        \node at (0.02,17pt)[anchor=north west, white]{\textbf{A}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.3\textwidth]
        {figures/crm_amir/progressions/step4-000017.png}
        \node at (0.02,17pt)[anchor=north west, white]{\textbf{B}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.3\textwidth]
        {figures/crm_amir/progressions/step4-000034.png}
            \node at (0.01, 0.01+17pt)[anchor=north west, white]{\textbf{C}};
    \end{tikzonimage}\\
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_amir/profiles-reduced/fit-comparison.pdf}%
        \node at (0.13, 0.88)[anchor=north west, black]{\textbf{B}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_amir/profiles/drag_force.pdf}%
        \node at (0.13, 0.88)[anchor=north west, black]{\textbf{D}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_amir/profiles/growth_rate.pdf}%
        \node at (0.13, 0.88)[anchor=north west, black]{\textbf{E}};
    \end{tikzonimage}\\
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_amir/profiles/rod_rigidity.pdf}%
        \node at (0.13, 0.88)[anchor=north west, black]{\textbf{F}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_amir/profiles/spring_tension.pdf}%
        \node at (0.13, 0.88)[anchor=north west, black]{\textbf{G}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_amir/profiles/damping.pdf}%
        \node at (0.13, 0.88)[anchor=north west, black]{\textbf{H}};
    \end{tikzonimage}%
    \caption{
        \textbf{TODO fix caption numbers and add missing stuff}
        (A) Microscopic image containing a single bacterium which is partially confined in a narrow
        tube on the bottom and free at the top where a controlled flow can be exerted on them.
        Division of the bacteria is inhibited.
        (B) A comparison of the initial and bent state show good visual agreement of our prediction
        with the experimental data.
        (C-G) Likelihood profiles for the individual parameters before and after performing two
        successive model reductions.
        The first step removes any damping effects by settings $\lambda=0$ while the second fixes
        the spring tension to its calculated optimum $\gamma=\gamma_\text{opt}$.
    }
    \label{fig:amir-bending-simulation}
\end{figure}

@@

In this section we study a collection of growing, dividing and physically interacting
\ac{ecoli}.
We use a series of microscopic images and corresponding segmentation masks for the individual
bacteria as the underlying dataset which is partially displayed in
Figure~\ref{fig:application-2-data-overview}.
With the previously developed extraction algorithm (see
section~\ref{subsec:position-extraction-algorithm}) we obtain the positions of our agents at all
time points.
These can be used as initial values and later within our parameter estimation routine.
We initialize our computational model (see section~\ref{sec:computational-model}) with the obtained
positions and optimize the parameters in order to find a set which best describes the observed
results.
The first case considers a time interval without cell-division while the second case does contain
division events.
Therefore, the first approach will optimize the direct comparison cost function (see
section~\ref{subsec:direct-metric}) while the latter approach has to account for division events
and thus requires the image-space metric method (see section~\ref{subsec:image-space-metric}).
These two scenarios present an opportunity to apply the previously developed methods within a
realistic context and validate our approach.

To show the usefulness of our previously developed methods
(see section~\ref{section:parameter-estimation}), we will use them to estimate the parameters of
individually growing \ac{ecoli} using a sequence of microscopic images.
The data we use~\cite{Trude1982} (see section~\ref{section:data-code-availability}) has been
automatically segmented with the omnipose~\cite{Cutler2022} algorithm to obtain the required cell
masks.

\textbf{Explain data more clearly. Refer to figure.}

>>>>

\begin{figure}[H]
    \centering
    \begin{tikzonimage}[width=0.3\textwidth]
        {figures/crm_amir/progressions/step4-000010.png}
        \node at (0.02,17pt)[anchor=north west, white]{\textbf{A}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.3\textwidth]
        {figures/crm_amir/progressions/step4-000017.png}
        \node at (0.02,17pt)[anchor=north west, white]{\textbf{B}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.3\textwidth]
        {figures/crm_amir/progressions/step4-000034.png}
            \node at (0.01, 0.01+17pt)[anchor=north west, white]{\textbf{C}};
    \end{tikzonimage}\\
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_amir/profiles-reduced/fit-comparison.pdf}%
        \node at (0.13, 0.88)[anchor=north west, black]{\textbf{B}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_amir/profiles/drag_force.pdf}%
        \node at (0.13, 0.88)[anchor=north west, black]{\textbf{D}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_amir/profiles/growth_rate.pdf}%
        \node at (0.13, 0.88)[anchor=north west, black]{\textbf{E}};
    \end{tikzonimage}\\
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_amir/profiles/rod_rigidity.pdf}%
        \node at (0.13, 0.88)[anchor=north west, black]{\textbf{F}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_amir/profiles/spring_tension.pdf}%
        \node at (0.13, 0.88)[anchor=north west, black]{\textbf{G}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_amir/profiles/damping.pdf}%
        \node at (0.13, 0.88)[anchor=north west, black]{\textbf{H}};
    \end{tikzonimage}%
    \caption{
        (A-C) Manually segmented experimental snapshots of by Amir et al.~\cite{Amir2014_2}.
        The red line within the green mask shows the extracted position.
        (A-C) Snapshots of bending experiment by Amir et al.~\cite{Amir2014_2} with overlayed
        manual segmentation mask (bright green) and extracted position (red).
        The sequence shows the rod initially (A), under flow (B) and after relaxation (C).
        (D) Comparison of simulated and experimental rod in bent and relaxed state.
        (D-H) Profile-likelihoods for the estimated parameters.
        Brigth blue curves represent the full model, dark blue a reduced model without damping
        $\lambda=0$ and orange curves when fixing the auxiliary spring tension parameter
        $\gamma=\gamma_\text{opt}$.
        With these reductions, every parameter is identifiable.
    }
    \label{fig:amir-bending-simulation}
\end{figure}

@@

In this section we study a collection of growing, dividing and physically interacting
\ac{ecoli}.
We consider two scenarios to apply the previously developed metrics within a realistic context and
validate our approach.
Our dataset~\cite{Trude1982} consists of of a time-series of phase-contrast images (see
Figure~\ref{fig:application-2-data-overview} (A)).
We segmented them using omnipose~\cite{Cutler2022} (see Figure~\ref{fig:application-2-data-overview}
(B)) and extracted their trajectories with the previously described pipeline
(Section~\ref{sec:supplement-extraction-algorithm}).
The model is initialized with the extracted position of the first snapshot and parameters are
optimized to minimize the respective metrics.

For our first case-study, we select 7 individual images spanning over an interval of
$\SI{20}{\minute}$ without any division events.
Figure~\ref{fig:application-2-data-overview} (C) displays the initial (left) an intermediate
(middle) and final state of this interval.
From these images we can already see that the cellular growth is the driving factor behind the
dynamics of the system.
We estimate the parameters of this systems via the direct position-space metric
(see Section~\ref{subsec:direct-metric}).

In the second case, we extend the previously used interval to 13 images over $\SI{85}{\minute}$
therefore including division events.
This change requires us to use the lineage-aware image-space metric (see
Section~\ref{subsec:image-space-metric}).
